Assessing \Htwo{} and \COtwo{} Network Strategies in Europe

### 2023-12-11 report/submissions/journal/paper/paper.tex
implement to-dos in Introduction, add source for Euro-Calliope

\begin{abstract}
    Hydrogen and carbon dioxide transportation are considered crucial in climate-neutral energy systems, with hydrogen enabling energy imports to high-demand areas and carbon transport aiding in emissions export from high-emission areas. Yet, possible synergies and competitions between the two systems are not fully understood. Our study employs optimization techniques to develop a cost-optimal European energy system, integrating transport, storage, and sequestration of both carbon dioxide and hydrogen, along with renewable energy sources. Results indicate that a hydrogen network is more cost-effective than a carbon network, facilitating low-cost hydrogen imports to Central Europe and local carbon capture and utilization. However, in a hybrid scenario, the carbon network effectively complements the hydrogen network, promoting carbon capture from biomass and reducing reliance on direct air capture. This research demonstrates the effectiveness of combining hydrogen and carbon transport networks with power grid expansion in achieving climate neutrality in Europe.
\end{abstract}

@@

The transition to a carbon-neutral European economy is a pressing challenge that demands coordinated action across various energy sectors. While management of both \COtwo{} and \Htwo{} is considered a critical component of this transition, a gap exists in understanding how new hydrogen infrastructures effectively interact with comprehensive carbon management technologies, including carbon capture, transport, use, and storage. Hydrogen offers an efficient way to transport and store energy. Carbon, on the other hand, can be effectively captured from industrial processes and the combustion of biomass, fossil fuels, and synthetic fuels through carbon capture (CC) techniques, or harvested from the atmosphere using direct air capture (DAC). Additionally, it can be stored in geological formations, a process known as carbon sequestration (CS). In combination, \COtwo{} and \Htwo{} networks build the basis for climate-neutral fuels needed for aviation, shipping and industrial feedstocks, and are central to carbon utilization (CU) strategies.

@@

However, up to this point, it remains unclear how the two transport systems of hydrogen and carbon may complement or replace each other, particularly when it comes to their transport systems. A hydrogen network facilitates the transport of energy from renewable sources to regions with geographically fixed hydrogen demand, such as for steelmaking, and also enables CU at the site of point-source capture. On the other hand, the carbon transport system allows for capturing and transporting carbon from areas with high emissions to regions with significant sequestration potential and abundant renewable sources, thereby enabling cost-effective CU.
In the literature, the two network approaches and underlying technologies have been discussed in a number of publications, all of which, however, dealt with the isolated effects~\cite{bakkenLinearModelsOptimization2008,morbeeOptimisedDeploymentEuropean2012,stewartFeasibilityEuropeanwideIntegrated2014,oeiModelingCarbonCapture2014,elahiMultiperiodLeastCost2014,burandtDecarbonizingChinaEnergy2019,middletonSimCCSOpensourceTool2020,bjerketvedtOptimalDesignCost2020,weiProposedGlobalLayout2021,damoreOptimalDesignEuropean2021,becattiniCarbonDioxideCapture2022}. Such techno-economic models, in contrast for example to Integrated Assessment Models, can account for the spatial distribution of carbon sources and sinks which are crucial provide a holistic view of the energy system and its technological interactions. The work in~\cite{neumannBenefitsHydrogenNetwork2022} examines the effect of a hydrogen network in Europe... (extend on hydrogen literature)
The publication by Morbee et al.~\cite{morbeeOptimisedDeploymentEuropean2012} optimizes the topology and capacity of a \COtwo{} network in Europe, but only considers the power sector without co-optimizing renewable deployment. This limited sectoral scope cannot capture important dynamics of carbon management, since it neglects the sectors that will need to handle most \COtwo{} in the future.
Another comprehensive example is found in~\cite{becattiniCarbonDioxideCapture2022}, which presents a mixed-integer model to optimize the time-evolution of a \COtwo{} transport system in Switzerland, connecting to a remote sequestration site in Norway. Hoewever, this limited spatial scope fails to consider other sequestration sites and co-benefits from connecting the \COtwo{} network to neighboring countries. To this end, it is often argued that \COtwo{} pipelines are a mature technology with an expected high learning rate, given the wide-spread installations in US and Canada for enhanced oil recovery~\cite{righettiSitingCarbonDioxide2017,friedmannNETZEROGEOSPHERICRETURN}.
% However, the models are often limited with regard to both geographical scope and detail. While representing a single country with spatial resolution may neglect synergies of international cooperation, a coarse grained representation of multiple countries may neglect important geographical properties.

>>>>

\begin{abstract}
    Hydrogen and carbon dioxide transportation are considered crucial in climate-neutral energy systems, with hydrogen transport enabling energy imports to high-demand areas and carbon transport aiding in emissions export from high-emission areas. Yet, possible synergies and competitions between the two systems are not fully understood. Our study employs optimization techniques to develop a cost-optimal European energy system, integrating transport, storage, and sequestration of both carbon dioxide and hydrogen, along with renewable energy sources. Results indicate that a hydrogen network is more cost-effective than a carbon network, facilitating low-cost hydrogen imports to Central Europe and local carbon capture and utilization. However, in a hybrid scenario, the carbon network effectively complements the hydrogen network, promoting carbon capture from biomass and reducing reliance on direct air capture. This research demonstrates the effectiveness of combining hydrogen and carbon transport networks with power grid expansion in achieving climate neutrality in Europe.
\end{abstract}

@@

The transition to a carbon-neutral European economy is a pressing challenge that demands coordinated action across various energy sectors. While management of both \COtwo{} and \Htwo{} is considered a critical component of this transition, a gap exists in understanding how new hydrogen infrastructure effectively interacts with carbon management technologies, including carbon capture, transport, storage, utilization, and sequestration. Hydrogen offers an efficient way to transport and store energy over long distances and timeframes, and can be produced from electrolysis or steam methane reforming (SMR). Carbon can be stored in geological formations to prevent it from entering the atmosphere, a process known as carbon sequestration (CS). Additionally, carbon is a crucial material input for carbonaceous fuels needed for aviation (synthetic kerosene from Fischer-Tropsch synthesis), shipping (synthetic methanol) and industrial feedstocks (e.g., synthetic methane). From hereon, we jointly refer to the production of these carbonaceous fuels as Carbon Utilization (CU). Carbon can be effectively captured from industrial processes and the combustion of biomass, fossil fuels, or synthetic carbonaceous fuels through point-source carbon capture (CC) techniques, or harvested from the atmosphere using direct air capture (DAC).

It is important to differentiate between the functions of carbon capture, carbon sequestration and carbon utilization. While both academic literature and common language mostly use the term ``carbon capture, utilization and storage'' (CCUS), or similar, to refer to all three functions, we argue that this is misleading. There is inherent benefit in separating these three functions geographically (linked through carbon transport), temporally (linked through carbon storage), and in terms of economic actors performing each function. CC should be performed at the place and with the technology that can capture carbon at the lowest cost (including a potential transport premium). CS should take place where sequestration potentials exist and carbon is available at the lowest cost, and CU where the combination of hydrogen and carbon inputs is available at lowest cost. By employing this separation in our model, we allow for a nuanced optimization and detailed analysis of the carbon management system.

@@

However, up to this point, it remains unclear how the two transport systems of hydrogen and carbon may complement or replace each other. Both can bridge the misalignment of sources (electrolysis, SMR, point-source CC, DAC) and sinks (CU, CS, industrial hydrogen demand) for carbon and hydrogen. A hydrogen network can supply regions with geographically fixed hydrogen demand, such as for steelmaking, with hydrogen from regions with the best renewable resources. A hydrogen network can also enable CU at the site of point-source CC. On the other hand, the carbon transport system allows for capturing and transporting carbon from areas with high emissions to regions with carbon sequestration potentials or high-quality renewable resources, the latter enabling cost-effective CU.
In the literature, the two network approaches and underlying technologies have been discussed in a number of publications, all of which, however, dealt with the isolated effects~\cite{bakkenLinearModelsOptimization2008, ,stewartFeasibilityEuropeanwideIntegrated2014,oeiModelingCarbonCapture2014,elahiMultiperiodLeastCost2014,burandtDecarbonizingChinaEnergy2019,middletonSimCCSOpensourceTool2020,bjerketvedtOptimalDesignCost2020,weiProposedGlobalLayout2021,damoreOptimalDesignEuropean2021,becattiniCarbonDioxideCapture2022}. Such techno-economic models, in contrast for example to Integrated Assessment Models, can account for the spatial distribution of carbon sources and sinks which are crucial provide a holistic view of the energy system and its technological interactions. The work in~\cite{neumannBenefitsHydrogenNetwork2022} examines the effect of a hydrogen network in Europe... (extend on hydrogen literature)

The publication by Morbee et al.~\cite{morbeeOptimisedDeploymentEuropean2012} optimizes the topology and capacity of a \COtwo{} network in Europe, but only considers the power sector without co-optimizing renewable deployment. This limited sectoral scope cannot capture important dynamics of carbon management, since it neglects the sectors that will need to handle most \COtwo{} in the future.
Another comprehensive example is found in~\cite{becattiniCarbonDioxideCapture2022}, which presents a mixed-integer model to optimize the time-evolution of a \COtwo{} transport system in Switzerland, connecting to a remote sequestration site in Norway. However, this limited spatial scope fails to consider other sequestration sites and co-benefits from connecting the \COtwo{} network to neighboring countries. Considering carbon pipeline as a technological option is a reasonable model assumption since \COtwo{} pipelines are a mature technology, given their wide-spread installations in the United States and Canada for enhanced oil recovery~\cite{righettiSitingCarbonDioxide2017,friedmannNETZEROGEOSPHERICRETURN}.
Other high-resolution energy system models do not feature detailed CU technologies and \COtwo{} transport. Euro-Calliope, for example, does neither feature carbon nor hydrogen transport, and thus forces generic synthetic fuel production to rely on captive electrolysis and DAC in the same region \cite{pickeringDiversityOptionsEliminate2022}.
In our model, we separate the functions of carbon capture (CC), carbon sequestration (CS) and utilization (CU).
% However, the models are often limited with regard to both geographical scope and detail. While representing a single country with spatial resolution may neglect synergies of international cooperation, a coarse grained representation of multiple countries may neglect important geographical properties.
